\section{Examples}\label{sec:examples}

We exemplarily solve the following problems in ASP:
$N$-Coloring (Section~\ref{subsec:ex:color}),
Traveling Salesperson (Section~\ref{subsec:ex:tsp}), and
Blocks-World Planning (Section~\ref{subsec:ex:block}).
While the first problem could likewise be solved within neighboring paradigms,
the second one requires checking reachability,
something that is rather hard to encode in either
Boolean Satisfiability~\cite{bihemawa08a} or
Constraint Programming~\cite{robewa06a}.
The third problem coming from the area of planning
illustrates incremental solving with \iclingo.
 

\subsection{$N$-Coloring}\label{subsec:ex:color}

As already mentioned in Section~\ref{subsec:asp},
it is custom in ASP to provide a \emph{uniform}
problem definition~\cite{martru99a,niemela99a,schlipf95a}.
We follow this methodology and separate the encoding
from an instance of the following problem:
given a (directed) graph, decide whether each node can be assigned
one of~$N$ colors such that any pair of adjacent nodes is colored differently.
Note that this problem is NP-complete for~$N\geq 3$
(see, e.g.,~\cite{papadimitriou94a}),
and thus it seems unlikely that a worst-case polynomial algorithm
can be found.
In view of this,
it is convenient to reduce the particular problem to
a declarative problem solving paradigm like ASP,
where efficient off-the-shelf tools like \gringo\ and \clasp\
are ready to solve the problem reasonably well.
Such a reduction is now exemplified.

\subsubsection{Problem Instance}\label{subsec:color:instance}
\input{figures/graph}

We consider directed graphs specified via facts over predicates
\pred{node}/$1$ and \pred{edge}/$2$.%
\footnote{%
  Directedness is not an issue in $N$-Coloring,
  but we will reuse our directed example graph in Section~\ref{subsec:ex:tsp}.}
The graph shown in Figure~\ref{fig:graph} is represented by the following set of facts:
%
\lstinputlisting{examples/graph.lp}
%
Recall from Section~\ref{subsec:lang:gringo} that~``\code{..}'' and~``\code{;}''
in the head expand to multiple rules, which are facts here.
Thus, the instance contains six nodes and 17 directed edges.

\subsubsection{Problem Encoding}\label{subsec:color:encoding}

We now proceed by encoding $N$-coloring via non-ground rules that are
independent of particular instances.
Typically, an encoding consists of a \emph{Generate}, a \emph{Define},
and a \emph{Test} part~\cite{lifschitz02a}.
As $N$-Coloring has a rather simple pattern, the following encoding does
not contain any Define part:
%
\lstinputlisting{examples/color.lp}
%
In Line~2, we use the \code{\#const} declarative,
described in Section~\ref{subsec:gringo:meta},
to install~\const{3} as default value for constant~\const{n} that is to be replaced
with the number~$N$ of colors.
(The default value can be overridden by invoking \gringo\ with option
 \code{--const n=$N$}.)
The Generate rule in Line~4 makes use of the \const{count} aggregate
(cf.\ Section~\ref{subsec:gringo:aggregate}).
For our example graph and~\const{1} substituted for~\var{X},
we obtain the following ground rule:%
\marginlabel{%
  The full ground program is obtained by invoking:\\
  \code{\mbox{~}gringo -t \textbackslash\\
    \mbox{~}examples/color.lp \textbackslash\\
    \mbox{~}examples/graph.lp}}
%
\begin{lstlisting}[numbers=none]
1 { color(1,1), color(1,2), color(1,3) } 1.
\end{lstlisting}
%
Note that~\code{\pred{node}(\const{1})} has been removed from the body,
as it is derived via a corresponding fact,
and similar ground instances are obtained for the other nodes~\const{2} to~\const{6}.
Furthermore, for each instance of \pred{edge}/2,
we obtain~\code{n} ground instances of the integrity constraint in Line~6,
prohibiting that the same color~\var{C} is assigned to the adjacent nodes.
Given~\code{\const{n}=\const{3}},
we get the following ground instances due to \code{\pred{edge}(\const{1},\const{2})}:
%
\begin{lstlisting}[numbers=none]
:- color(1,1), color(2,1).
:- color(1,2), color(2,2).
:- color(1,3), color(2,3).
\end{lstlisting}
%
Again note that \code{\pred{edge}(\const{1},\const{2})},
derived via a fact, has been removed from the body.

\subsubsection{Problem Solution}\label{subsec:color:solution}
\input{figures/color}

Provided that a given graph is colorable with~\const{n} colors,
a solution can be read off an answer set of the program consisting
of the instance and the encoding.
For the graph in Figure~\ref{fig:graph},
the following answer set can be computed:%
\marginlabel{%
  To find an answer set, invoke:\\
  \code{\mbox{~}gringo \textbackslash\\
        \mbox{~}examples/color.lp \textbackslash\\
        \mbox{~}examples/graph.lp |\rlap{\:\textbackslash}\\
        \mbox{~}clasp}\\
  or alternatively:\\
  \code{\mbox{~}clingo \textbackslash\\
        \mbox{~}examples/color.lp \textbackslash\\
        \mbox{~}examples/graph.lp}}
%
\begin{lstlisting}[numbers=none]
Answer: 1
... color(1,2) color(2,1) color(3,1) color(4,3) color(5,2) color(6,3)
\end{lstlisting}
%
Note that we have omitted the six instances of \pred{node}/$1$ and the
17 instances of \pred{edge}/$2$ in order to emphasize the actual solution,
which is depicted in Figure~\ref{fig:color}.
Such output projection can also be specified within a logic program file by
using the declaratives \code{\#hide} and \code{\#show},
described in Section~\ref{subsec:gringo:meta}.


\subsection{Traveling Salesperson}\label{subsec:ex:tsp}

We now consider the well-known Traveling Salesperson Problem (TSP),
where the task is to decide whether there is a round trip that visits
each node in a graph exactly once (viz., a Hamiltonian cycle) and whose
accumulated edge costs must not exceed some budget~$B$.
We tackle a slightly more general variant of the problem by not
a priori fixing~$B$ to any integer.
Rather,
we want to compute a minimum budget~$B$ along with a round trip of cost~$B$.
This problem is FP$^\textrm{NP}$-complete (cf.~\cite{papadimitriou94a}),
that is, it can be solved with a polynomial number of queries to an NP-oracle.
As with $N$-Coloring,
we provide a uniform problem definition by separating the encoding from instances.

\subsubsection{Problem Instance}\label{subsec:tsp:instance}
\input{figures/costs}

We reuse graph specifications in terms of predicates \pred{node}/$1$ and \pred{edge}/$2$
as in Section~\ref{subsec:color:instance}.
In addition, facts over predicate \pred{cost}/$3$ are used to define edge costs:
%
\lstinputlisting{examples/costs.lp}
%
Figure~\ref{fig:costs} shows the (directed) graph from Figure~\ref{fig:graph}
along with the associated edge costs.
Symmetric edges have the same costs here,
but differing costs would also be possible.


\subsubsection{Problem Encoding}\label{subsec:tsp:encoding}

The first subproblem consists of describing a Hamiltonian cycle,
constituting a candidate for a minimum-cost round trip.
Using the Generate-Define-Test pattern~\cite{lifschitz02a},
we encode this subproblem via the following non-ground rules:
%
\lstinputlisting{examples/ham.lp}
%
The Generate rules in Line~2 and~3 assert that every node must have
exactly one outgoing and exactly one incoming edge, respectively,
belonging to the cycle.
By inserting the available edges, for node~\const{1},
Line~2 and~3 are grounded as follows:%
\marginlabel{%
  The full ground program is obtained by invoking:\\
  \code{\mbox{~}gringo -t \textbackslash\\
    \mbox{~}examples/ham.lp \textbackslash\\
    \mbox{~}examples/min.lp \textbackslash\\
    \mbox{~}examples/costs.lp \textbackslash\\
    \mbox{~}examples/graph.lp}}
%
\begin{lstlisting}[numbers=none]
1 { cycle(1,2), cycle(1,3), cycle(1,4) } 1.
1 { cycle(3,1), cycle(4,1) } 1.
\end{lstlisting}
%
Observe that the first rule groups all outgoing edges of node~\const{1},
while the second one does the same for incoming edges.
We proceed by considering the Define rules in Line~5 and~6,
which recursively check whether nodes are reached by a cycle candidate
produced via the Generate part.
Note that the rule in Line~5 builds on the assumption that the cycle
``starts'' at node~\code{1}, that is,
any successor~\var{Y} of~\code{1} is reached by the cycle.
The second rule in Line~6 states that, from a reached node~\var{X},
an adjacent node~\var{Y} can be reached via a further edge in the cycle.
Note that this definition admits positive recursion
among the ground instances of \pred{reached}/$1$,
in which case a ground program is called \emph{non-tight}~\cite{erdlif03a,fages94a}.
The ``correct'' treatment of (positive) recursion due to the justification
requirement w.r.t.\ the reduct (cf.\ Section~\ref{subsec:semantics})
is a particular feature of answer set semantics,
which is hard to mimic in either
Boolean Satisfiability~\cite{bihemawa08a} or
Constraint Programming~\cite{robewa06a}.
In our present problem, this feature makes sure that all nodes are reached
by a global cycle from node~\code{1}, thus, excluding isolated subcycles.
In fact, the Test in Line~8 stipulates that every node in the given graph
is reached, that is, the instances of \pred{cycle}/$2$ in an answer set
must be the edges of a Hamiltonian cycle.%
\marginlabel{%
  To compute the six Hamiltonian cycles
  for the graph in Figure~\ref{fig:graph}, invoke:\\
  \code{\mbox{~}gringo \textbackslash\\
        \mbox{~}examples/ham.lp \textbackslash\\
        \mbox{~}examples/graph.lp |\rlap{\:\textbackslash}\\
        \mbox{~}clasp -n 0}\\
  or alternatively:\\
  \code{\mbox{~}clingo -n 0 \textbackslash\\
        \mbox{~}examples/ham.lp \textbackslash\\
        \mbox{~}examples/graph.lp}}
Finally, the additional Display part in Line~10 and~11 states that answer
sets should be projected to instances of \pred{cycle}/$2$, as only they
describe a solution.
We have so far not considered edge costs,
and answer sets for the above 
part of the encoding correspond to Hamiltonian cycles,
that is, candidates for a minimum-cost round trip.

In order to minimize costs,
we add the following optimization statement:
%
\lstinputlisting[firstnumber=13]{examples/min.lp}
%
Here, edges belonging to the cycle are weighted according to their costs.
After grounding, the minimization in Line~14 ranges over~17 instances
of \pred{cycle}/$2$, one for each weighted edge in Figure~\ref{fig:costs}.

\subsubsection{Problem Solution}\label{subsec:tsp:solution}
\input{figures/tsp}

Finally, we explain how the unique minimum-cost round trip
(depicted in Figure~\ref{fig:tsp}) can be computed.
The catch is that we are now interested in optimal answer sets,
rather than in arbitrary ones.
In order to determine the optimum, we can start by gradually
decreasing the costs associated to answer sets
until we cannot find a strictly better one anymore.
To this end, it is important to invoke \clasp\ (or \clingo)
with option ``\code{-n~0}'' (cf.\ Section~\ref{subsec:opt:clasp}),
making it enumerate successively better answer sets
w.r.t.\ the provided optimization statements (cf.\ Section~\ref{subsec:gringo:optimize}).
Any answer set is printed as soon as it has been computed,
and the last one is optimal.
If there are multiple optimal answer sets, an arbitrary one among them is computed.
For the graph in Figure~\ref{fig:costs},
the optimal answer set (cf.\ Figure~\ref{fig:tsp}) is unique,
and its computation can proceed as follows:%
\marginlabel{%
  To compute the minimum-cost round trip
  for the graph in Figure~\ref{fig:costs}, invoke:\\
  \code{\mbox{~}gringo \textbackslash\\
    \mbox{~}examples/ham.lp \textbackslash\\
    \mbox{~}examples/min.lp \textbackslash\\
    \mbox{~}examples/costs.lp \textbackslash\\
    \mbox{~}examples/graph.lp |\rlap{\:\textbackslash}\\
    \mbox{~}clasp -n 0}\\
  or alternatively:\\
  \code{\mbox{~}clingo -n 0 \textbackslash\\
    \mbox{~}examples/ham.lp \textbackslash\\
    \mbox{~}examples/min.lp \textbackslash\\
    \mbox{~}examples/costs.lp \textbackslash\\
    \mbox{~}examples/graph.lp}}
%
\begin{lstlisting}[numbers=none]
Answer: 1
cycle(1,3) cycle(2,4) cycle(3,5) \
cycle(4,1) cycle(5,6) cycle(6,2)
Optimization: 13
Answer: 2
cycle(1,2) cycle(2,5) cycle(3,4) \
cycle(4,1) cycle(5,6) cycle(6,3)
Optimization: 11
\end{lstlisting}
%
Given that no answer is obtained after the second one,
we know that~\code{11} is the optimum value,
but there might be more optimal answer sets that have not been computed yet.
In order to find them too, we can use the following command line options
of \clasp\ (cf.\ Section~\ref{subsec:opt:clasp}):
``\code{--opt-value=11}'' in order to initialize the optimum and
``\code{--opt-all}'' to compute also equitable
(rather than only strictly better) answer sets.%
\marginlabel{%
  The full invocation is:\\
  \code{\mbox{~}gringo \textbackslash\\
    \mbox{~}examples/ham.lp \textbackslash\\
    \mbox{~}examples/min.lp \textbackslash\\
    \mbox{~}examples/costs.lp \textbackslash\\
    \mbox{~}examples/graph.lp |\rlap{\:\textbackslash}\\
    \mbox{~}clasp -n 0 \textbackslash\\
    \mbox{~}--opt-value=11 \textbackslash\\
    \mbox{~}--opt-all}\\
  or alternatively:\\
  \code{\mbox{~}clingo -n 0 \textbackslash\\
    \mbox{~}--opt-value=11 \textbackslash\\
    \mbox{~}--opt-all \textbackslash\\
    \mbox{~}examples/ham.lp \textbackslash\\
    \mbox{~}examples/min.lp \textbackslash\\
    \mbox{~}examples/costs.lp \textbackslash\\
    \mbox{~}examples/graph.lp}}
After obtaining only the second answer given above,
we are sure that this is the unique optimal answer set,
whose associated edge costs (cf.\ Figure~\ref{fig:tsp}) correspond to
the reported optimization value~\code{11}.
Note that, with \const{maximize} statements in the input, this correlation
might be less straightforward because they are compiled into \const{minimize}
statements in the process of generating \lparse's output format~\cite{lparseManual}.
Furthermore, if there are multiple optimization statements,
\clasp\ (or \clingo) will report separate optimization values ordered by significance.
Finally, the two-stage invocation scheme exercised above,
first determining the optimum and afterwards all (and only) optimal answer sets,
is recommended for general use.
Otherwise,
if using option ``\code{--opt-all}'' right away without knowing the optimum,
one risks the enumeration of plenty suboptimal answer sets.
We also invite everyone to explore command line option ``\code{--opt-restart}''
(cf.\ Section~\ref{subsec:opt:clasp})
in order to see whether it improves search efficiency on optimization problems.


\subsection{Blocks-World Planning}\label{subsec:ex:block}

The Blocks-World is a well-known planning domain,
where finding \emph{shortest} plans has received particular attention~\cite{gupnau92a}.
In an eventually propositional formalism like ASP,
a bound on the plan length must be fixed before search can proceed.
This is usually accomplished by including some constant~\const{t}
in an encoding, which is then replaced with the actual bound during grounding.
Of course, if the length of a shortest plan is unknown,
an ASP system must repeatedly be queried while varying the bound.
With a traditional ASP system, processing
the same planning problem with a different bound
involves grounding and solving from scratch.
In order to reduce such redundancies,
the incremental ASP system \iclingo~\cite{gekakaosscth08a}
can gradually increase a bound, % thereby,
doing only the necessary additions in each step.
Note that planning is a natural application domain for \iclingo,
but other problems including some mutable bound can be addressed too, thus,
\iclingo\ is not a specialized planning system.
However, we use Blocks-World Planning to illustrate the exploitation of
\iclingo's incremental computation mode.


\subsubsection{Problem Instance}\label{subsec:block:instance}

As with the other two problems above,
an instance is given by a set of facts, here,
over predicates \pred{block}/$1$ (declaring blocks),
\pred{init}/$1$ (defining the initial state), and
\pred{goal}/$1$ (specifying the goal state).
A well-known Blocks-World instance is described by:%
\footnote{%
  Blocks-World instances ``\code{examples/world$i$.lp}'' for
  $i\in\{\code{0},\code{1},\code{2},\code{3},\code{4}\}$
  are adaptations of the instances provided at~\cite{erdemBW}.}
%
\lstinputlisting{examples/world0.lp}
%
Note that the facts in Line~13--15 and~24--26 specify the initial
and the goal state depicted in Line~9-11 and~19--22, respectively.
We here use (uninterpreted) function \const{on}/$2$ to illustrate another
important feature available in \gringo, \clingo, and~\iclingo, namely,
the possibility of instantiating variables to compound terms.


\subsubsection{Problem Encoding}\label{subsec:block:encoding}

Our Blocks-World Planning encoding for \iclingo\ makes use of declaratives
\const{\#base}, \const{\#cumulative}, and \const{\#volatile},
separating the encoding into a static, a cumulative, and a volatile (query) part.
Each of them can be further refined into Generate, Define, Test, and Display constituents, 
as indicated in the comments below:
%
\lstinputlisting{examples/blocks.lp}
%
In the initial \const{\#base} part (Line~1--5),
we define blocks and constant \const{table}
as instances of predicate \pred{location}/$1$.
Moreover, we use instances of \pred{init}/$1$
to initialize predicate \pred{holds}/$2$ for step~\const{0}, thus,
defining the conditions before the first incremental step.
Note that variable~\var{F} is instantiated to compound terms
over function \const{on}/$2$.

The \const{\#cumulative} statement in Line~7 declares constant~\code{t}
as a placeholder for step numbers in the cumulative encoding part below.
Here, the Generate rule in Line~9 states that exactly one block~\var{X}
must be moved to a location~\var{Y} (different from~\var{X}) at each step~\code{t}.
The integrity constraint in Line~11--13 is used to test whether
moving block~\var{X} to location~\var{Y} is possible at step~\code{t}
by denying~\code{\pred{move}(X,Y,t)} to hold if there
is either some block~\var{A} on~\var{X} or some block~\var{B} distinct from~\var{X} on~\var{Y}
(this condition is only checked if~\var{Y} is a block, viz., different from constant \const{table})
at step~\code{t-1}.
Finally, the Define rule in Line~15 propagates
a move to the state at step~\code{t},
while the rule in Line~16--17 states that a block~\var{X} stays on a location~\var{Z}
if it is not moved to any other location~\var{Y}.
Note that we require atoms \code{\pred{block}(\var{X})} and \code{\pred{location}(\var{Z})}
to bind variables~\var{X} and~\var{Z} in the body of the latter rule,
as recursive predicate \pred{holds}/$2$ cannot be used for this purpose
(cf.\ the definition of level-restrictedness in Section~\ref{subsec:lambda}).

The \const{\#volatile} statement in Line~19 declares the next part as a
query depending on step number~\const{t}, but not accumulated over
successive steps.
In fact, the integrity constraint in Line~21 tests whether goal conditions are
satisfied at step~\const{t}.

Our incremental encoding concludes with a second \const{\#base} part,
as specified in Line~23.
Note that, for the meta-statements with Display functionality (Line~25--26),
it is actually unimportant whether they belong to a static, cumulative, or
volatile program part, as answer sets are projected
(to instances of \pred{move}/$3$) in either case.
However, by ending the encoding file with a \const{\#base} statement,
we make sure that the contents of a concatenated instance file
is included in the static program part.
This is also the default of \iclingo\
(as well as of \gringo\ and \clingo\ that can be used for non-incremental computations),
but applying this default behavior triggers a warning (cf.\ Section~\ref{subsec:warn}).%
\marginlabel{%
  To observe the ground program dealt with internally \iclingo\
  at a step $n$, invoke:\\
  \code{\mbox{~}gringo -t \textbackslash\\
        \mbox{~}--ifixed=$n$ \textbackslash\\
        \mbox{~}examples/blocks.lp \textbackslash\\
        \mbox{~}examples/world$i$.lp}
  where $i\in\{\code{0},\code{1},\code{2},\code{3},\code{4}\}$.}

Finally, let us stress important prerequisites for obtaining
a well-defined incremental computation result from \iclingo.
First, the ground instances of head atoms of rules in the
static, cumulative, and volatile program part must be pairwisely disjoint.
Furthermore, ground instances of head atoms in the volatile part
must not occur in the static and cumulative parts,
and those of the cumulative part must not be used in the static part.
Finally, ground instances of head atoms in either the cumulative or the volatile part
must be different for each pair of distinct steps.
This is the case for our encoding because both atoms over \pred{move}/$3$
and those over \pred{holds}/$2$ include~\const{t} as an argument in the
heads of the rules in Line~9, 15, and~16--17.
As the smallest step number to replace~\const{t} with is~\const{1},
there also is no clash with the ground atoms over \pred{holds}/$2$
obtained from the head of the static rule in Line~5.
Further details on the sketched requirements and their formal background can
be found in~\cite{gekakaosscth08a}.
Arguably, many problems including some mutable bound can be encoded
such that the prerequisites apply.
Some attention should of course be spent on
putting rules into the right program parts.


\subsubsection{Problem Solution}\label{subsec:block:solution}

We can now use \iclingo\ to \emph{incrementally} compute the shortest
sequence of moves that brings us from the initial to the goal state
depicted in the instance in Section~\ref{subsec:block:instance}:%
\marginlabel{%
  To this end, invoke:\\
  \code{\mbox{~}iclingo -n 0 \textbackslash\\
        \mbox{~}examples/blocks.lp \textbackslash\\
        \mbox{~}examples/world0.lp}}
%
\begin{lstlisting}[numbers=none]
Answer: 1
move(b2,table,1) move(b1,b0,2) move(b2,b1,3)
\end{lstlisting}
%
This unique answer set tells us that the given problem instance can
be solved by moving block~\const{b2} to the~\const{table} in order to
then put~\const{b1} on top of~\const{b0} and finally~\const{b2} on top of~\const{b1}.
This solution is computed by \iclingo\ in three grounding and solving steps,
where, starting from the \const{\#base} program, constant~\const{t}
is successively replaced with step numbers~\const{1}, \const{2}, and~\const{3}
in the~\const{\#cumulative} and in the~\const{\#volatile} part.
While the query postulated in the~\const{\#volatile} part cannot be
fulfilled in steps~\const{1} and~\const{2}, \iclingo\ 
stops its incremental computation after finding an answer set in step~\const{3}.
The scheme of iterating steps until finding at least one answer set
is the default behavior of~\iclingo,
which can be customized via command line options
(cf.\ Section~\ref{subsec:opt:iclingo}).

Finally, let us describe how solutions obtained via an incremental computation
can be computed in the standard way, that is, in a single pass.
To this end, the step number can be fixed to some~$n$ via
option ``\code{--ifixed=$n$}'' (cf.\ Section~\ref{subsec:opt:gringo}),
\marginlabel{%
  For non-incremental solving, invoke:\\
  \code{\mbox{~}gringo --ifixed=$n$ \textbackslash\\
        \mbox{~}examples/blocks.lp \textbackslash\\
        \mbox{~}examples/world$i$.lp |\rlap{\textbackslash}\\
        \mbox{~}clasp -n 0}\\
  or alternatively:\\
  \code{\mbox{~}clingo -n 0 \textbackslash\\
        \mbox{~}--ifixed=$n$ \textbackslash\\
        \mbox{~}examples/blocks.lp \textbackslash\\
        \mbox{~}examples/world$i$.lp}
  where $i\in\{\code{0},\code{1},\code{2},\code{3},\code{4}\}$.}
enabling \gringo\ or \clingo\ to generate the ground program present
inside \iclingo\ at step~$n$.
Note that \const{\#volatile} parts are here only instantiated for the final step~$n$,
while \const{\#cumulative} rules are added for all steps $\const{1},\dots,n$.
Option ``\code{--ifixed=$n$}'' can be useful for investigating the contents
of a ground program dealt with at step~$n$ or for using an external solver
(other than \clasp).
In the latter case, repeated invocations with varying~$n$
are required if the bound of an optimal solution is a priori unknown.


%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "guide"
%%% End: 
